---
title: "Gov 50 Final Project"
author: "Dominic Sheeks"
description: "My Final Project"
output:
  distill::distill_article:
    self_contained: false
---



## Project Proposal

In the United States, has the increased usage of narcotic analgesics – like fentanyl – decreased the number of heroin-related overdose fatalities? In this project, I will compare the by-the-year numbers (from 1999 to 2020) of overdose deaths caused by each of these opioids, making note of any significant correlations between the two. It is my prediction that as the availability of fentanyl and other synthetic analgesics increased, the use of heroin and other opioids will have decreased, thus resulting in lower fatal overdoses due to heroin. My dataset, provided by the CDC – coded as ‘overdoseRates’ – spans over the aforementioned years and analyzes the U.S. as a whole. The main variable of interest is deaths caused by synthetic opioid analgesics (coded as ‘synth_opioid_analgesics’), with the independent variable in question being fatal heroin overdoses (coded as ‘heroin’). Both variables operate on a percent-per-100k-people basis: “1” being one percent of said people, and “20” meaning twenty percent of 100,000 people. To compare the correlation between the two variables, I will have to look at variable ‘Year’. If the variable ‘Year’ has a positive correlation with the variable ‘synth_opiod_analgesics’  but a negative correlation with ‘heroin’, this would support my hypothesis. However, there may be other factors that contribute to the increase or decrease in the percent of overdose fatalities with regard to a given opioid.


```{r}
library(googlesheets4)

overdoseRates <- read_sheet("https://docs.google.com/spreadsheets/d/135BpS8CTn1LdcUvf7ZwK6GIKSt-LyIPJ_uqbSWqwYgo/edit#gid=0")

overdoseRates
```

```{r}
library(ggplot2)

milestone_3 <- ggplot(data = overdoseRates,
       mapping = aes(x = Year,
                     y = synth_opioid_analgesics)) +
  geom_point(color = "blue") +
  geom_smooth() +
  labs(x = "Year",
       y = "Death Rate by Synth Opioids - Fetanyl")

milestone_3

```


## First Visualization for Milestone 3 